# Applying Covariate Adjustment: ANCOVA, G-Computation, DR-WLS

In this next section, we will use R to produce the ANCOVA estimator, get an estimate of its standard error, and compare its precision to an unadjusted analysis using R. For this example we will use the Buprenorphine tapering trial data (CTN-03).

```{r Load-CTN-03-Data, eval = FALSE}
data_url <-
  "https://github.com/jbetz-jhu/CovariateAdjustmentTutorial/raw/main/SIMULATED_CTN03_220506.Rdata"

load(file = url(data_url))
```

This .Rdata file contains two datasets: `ctn03_sim`, which has no missing data, and `ctn03_sim_mar`, where a simulated missing data mechanism has been applied.


```{r ANCOVA-CTN03-Step-1-Fit-Model}
# 1. Regress final outcome (VAS at end of taper) on treatment assignment and
# outcome assessed at baseline (VAS at baseline)
vas_ancova_1 <-
  lm(
    formula = vas_crave_opiates_eot ~ arm + vas_crave_opiates_bl,
    data = ctn03_sim_mar
  )

summary(vas_ancova_1)
```


Note that the standard error reported by `lm()` is the model-based standard error, not a robust standard error or bootstrap standard error.


```{r ANCOVA-CTN03-Step-2-Predict-Outcomes}
# 2. Generate predictions based on fitted model:
# Their predicted outcome if they were assigned to treatment (7-day taper)
# Their predicted outcome if they were assigned to control (28-day taper)
expected_vas_7_day <-
  predict(
    object = vas_ancova_1,
    newdata =
      within(
        data = ctn03_sim_mar,
        expr = {arm = "7-day"}
      ),
    type = "response"
  )

expected_vas_28_day <-
  predict(
    object = vas_ancova_1,
    newdata =
      within(
        data = ctn03_sim_mar,
        expr = {arm = "28-day"}
      ),
    type = "response"
  )


data.frame(
  expected_vas_7_day,
  expected_vas_28_day,
  ctn03_sim_mar[, c("arm", "vas_crave_opiates_eot", "vas_crave_opiates_bl")]
) %>% 
  head()
```


Once we've generated these predictions, we can compute the ANCOVA estimate of the average treatment effect:


```{r ANCOVA-CTN03-Step-3-Compute-Estimand}
mean(expected_vas_7_day)
mean(expected_vas_28_day)
vas_ancova_estimate <- mean(expected_vas_7_day) - mean(expected_vas_28_day)
vas_ancova_estimate
```


Notice this is the exact same as the regression coefficient for `arm` in the regression model, since there's no treatment-covariate interactions and an identity link function. We can obtain the standard error of the estimate two ways. The first way is using the `margins::margins()` command, using the robust standard errors from `sandwich::vcovHC`:

```{r ANCOVA-CTN03-Compute-SE-Margins}
vas_ancova_margins <-
  margins::margins(
    model = vas_ancova_1,
    # Specify treatment variable
    variables = "arm",
    # Convert to outcome scale, not link scale
    type = "response",
    # Obtain robust standard errors
    vcov = sandwich::vcovHC(x = vas_ancova_1, type = "HC3")
  )

summary(object = vas_ancova_margins, level = 0.95)
```


You'll see that we now have a standard error, p-value under the hypothesis that the marginal effect is 0, and a 95% Confidence Interval for the estimate. Another way to obtain these is the bias corrected and accelerated (BCa) non-parametric bootstrap:


```{r ANCOVA-CTN03-Compute-SE-Boot, eval = FALSE}
# Write a function to produce the ANCOVA estimate
margins_fun <-
  function(
    data,
    indices = NULL,
    formula,
    family,
    term,
    contrast = c("difference", "ratio")[1]
  ){
    # Input data must be a data.frame
    if(!all(class(data) == "data.frame")){
      stop("`data` must be a data.frame: use `as.data.frame()` for a tibble.")
    }
    
    # If bootstrap indices not supplied, use entire dataset
    if(is.null(indices)) indices <- 1:nrow(data)
    
    data <- data[indices,]
    
    glm_fit <-
      glm(
        formula = formula,
        family = family,
        data = data
      )
    
    tx_levels <- levels(data[, term])
    
    e_y_1 <-
      predict(
        object = glm_fit,
        newdata = 
          within(
            data,
            expr = assign(x = term, value = tx_levels[2])
          ),
        
        type = "response"
      )
    
    e_y_0 <-
      predict(
        object = glm_fit,
        newdata = 
          within(
            data,
            expr = assign(x = term, value = tx_levels[1])
          ),
        
        type = "response"
      )
    
    if(contrast == "difference"){
      return(mean(e_y_1) - mean(e_y_0))
    } else if (contrast == "ratio"){
      return(mean(e_y_1)/mean(e_y_0))
    }
  }

vas_ancova_boot <-
  boot::boot(
    data = ctn03_sim_mar,
    statistic = margins_fun,
    R = 10000,
      formula = vas_crave_opiates_eot ~ arm + vas_crave_opiates_bl,
    family = gaussian(link = "identity"),
    term = "arm"
  )
```

```{r ANCOVA-CTN03-Report-SE-Boot}
# Bootstrap results
vas_ancova_boot

# Bootstrap Standard Error
sd(vas_ancova_boot$t[,1])
```


We can extract the standard error and produce a 95% confidence interval:


```{r ANCOVA-CTN03-Compute-CI-Boot, eval = FALSE}
vas_ancova_boot_ci <-
  boot::boot.ci(
    boot.out = vas_ancova_boot,
    conf = 0.95,
    type = "bca"
  )
```


```{r ANCOVA-CTN03-Report-CI-Boot, eval = FALSE}
vas_ancova_boot_ci
```


We can compare these to the results from `margins::margins()`:



```{r Compare-Margins-Boot}
# Margins SE
ancova_margins_se <- summary(object = vas_ancova_margins, level = 0.95)$SE
ancova_margins_se

# Bootstrap SE
ancova_boot_se <- sd(vas_ancova_boot$t[,1])
ancova_boot_se

# Compare as ratio
ancova_margins_se/ancova_boot_se
```


The standard errors are nearly identical using the estimated marginal effect with robust standard errors or the bootstrap.




--------------------------------------------------------------------------------




## Calculating Precision Gain

Asymptotically, the ANCOVA estimate should have precision that is equal or better than an unadjusted analysis. We can calculate the precision gain by taking the ratio of the variances, i.e. the squared standard errors, of the adjusted estimator to the unadjusted estimator.

```{r Compute-Unadjusted-Estimator}
vas_t_test <-
  t.test(
    formula = vas_crave_opiates_eot ~ arm,
    data = ctn03_sim_mar,
    var.equal = FALSE
  )

vas_t_test

# Get unadjusted standard error
t_test_se <- vas_t_test$stderr
t_test_se
```


The precision gain is equal to the ratio of the variance, which is the square of the standard error:


```{r Compute-Ratio-Variances}
# Percentage reduction in variance adjusting for baseline outcome
100*(1 - (ancova_margins_se/t_test_se)^2)
```


The precision gain in this particular example is rather small, but we have only adjusted for one potential covariate. However, there are several other baseline variables we did not include in the model. We can see if the gain in precision is larger when these other variables are included. We should also include the stability dose in the outcome, since randomization was stratified by this variable:


```{r ANCOVA-CTN03-Larger-Model}
vas_ancova_2 <-
  glm(
    formula = vas_crave_opiates_eot ~ 
      arm + vas_crave_opiates_bl + stability_dose +
      arsw_score_bl + cows_total_score_bl +
      vas_current_withdrawal_bl + vas_study_tx_help_bl + uds_any_positive_bl,
    data = ctn03_sim_mar
  )

vas_ancova_margins_2 <-
  margins::margins(
    model = vas_ancova_2,
    # Specify treatment variable
    variables = "arm",
    # Convert to outcome scale, not link scale
    type = "response",
    # Obtain robust standard errors
    vcov = sandwich::vcovHC(x = vas_ancova_2, type = "HC3")
  )

ancova_margins_2_se <- summary(object = vas_ancova_margins_2, level = 0.95)$SE

# Percentage reduction in variance adjusting for baseline outcome
100*(1 - (ancova_margins_2_se/t_test_se)^2)
```


Including these other covariates leads to a larger gain in precision.




--------------------------------------------------------------------------------


## Covariate Adjustment with Binary Outcomes

When the outcome is binary, the only difference in the approach is fitting a logistic regression with a nonlinear link instead of multiple regression with an identity link:

```{r Model-Fit-CTN03-UDS}
uds_opioids_glm <-
  glm(
    formula = 
      uds_opioids_eot == "Negative" ~ 
      arm + uds_opioids_bl + stability_dose +
      arsw_score_bl + cows_total_score_bl +
      vas_current_withdrawal_bl + vas_study_tx_help_bl,
    data = ctn03_sim_mar,
    family = binomial(link = "logit")
  )

summary(uds_opioids_glm)
```


Logistic regression gives coefficients on the log odds scale, which can be exponentiated to obtain a conditional odds ratio, a relative measure. The `margins::margins` function gives the marginal estimate on the risk difference scale, which is an absolute measure:


```{r G-Computation-CTN03-UDS}
uds_opioids_margins <-
  margins::margins(
    model = uds_opioids_glm,
    # Specify treatment variable
    variables = "arm",
    # Convert to outcome scale, not link scale
    type = "response",
    # Obtain robust standard errors
    vcov = sandwich::vcovHC(x = uds_opioids_glm, type = "HC3")
  )

summary(uds_opioids_margins)

confint(uds_opioids_margins)
```

A risk ratio could be obtained by using `margins_fun` from above, specifying `contrast = "ratio"`, and obtaining the standard error and confidence interval via the bootstrap. In the event that a marginal odds ratio is of interest, specifying `type = "link"` in the `margins::margins` call gives a marginal log odds ratio, which could be exponentiated to obtain a marginal odds ratio.




--------------------------------------------------------------------------------


## Stratified Randomization & Covariate Adjustment

When stratified randomization is employed, additional benefits in precision can be realized by using covariate adjustment and a variance estimator designed for covariate adaptive designs. First, source the ICAD code from Github:


```{r Load-ICAD}
icad_link <-
  "https://raw.githubusercontent.com/BingkaiWang/covariate-adaptive/master/R/ICAD.R"
source(url(icad_link))
```


Next, we specify which covariates are included in the model, excluding those in the strata variable, and call ICAD:


```{r DR-WLS-CTN03-VAS}
baseline_covariates <-
  c("uds_opioids_bl", "arsw_score_bl", "cows_total_score_bl",
    "vas_current_withdrawal_bl", "vas_study_tx_help_bl")

vas_icad <-
  ICAD(
    Y = 1*(ctn03_sim_mar$uds_opioids_eot == "Negative"), 
    A = 1*(ctn03_sim_mar$arm == "7-day"), # Treatment indicator - Must be 1/0
    Strata = ctn03_sim_mar$stability_dose, # Randomization Stratum
    W = ctn03_sim_mar[, baseline_covariates], # Baseline Covariates
    pi = 0.5, # 1:1 Randomization,
    family = "binomial"
  )

vas_icad
```

We can see that the 'adjusted-for-all' estimator, which uses a specialized variance estimator for covariate adaptive designs, has a lower variance than the covariate adjusted estimator obtained using `margins`, even though we are adjusting for the same covariates. This is the result of the additional precision gains from the covariate-adaptive variance estimator. The doubly robust weighted least squares (`drwls`) estimator will be discussed later on.

Specifying `family = "binomial"` allows for marginal effects with binary outcomes:

```{r DR-WLS-CTN03-UDS}
baseline_covariates <-
  c("vas_crave_opiates_bl", "arsw_score_bl",
    "cows_total_score_bl", "vas_current_withdrawal_bl", "vas_study_tx_help_bl",
    "uds_any_positive_bl")

uds_icad <-
  ICAD(
    Y = ctn03_sim_mar$vas_crave_opiates_eot, 
    A = 1*(ctn03_sim_mar$arm == "7-day"), # Treatment indicator - Must be 1/0
    Strata = ctn03_sim_mar$stability_dose, # Randomization Stratum
    W = ctn03_sim_mar[, baseline_covariates], # Baseline Covariates
    pi = 0.5 # 1:1 Randomization
  )

uds_icad
```




--------------------------------------------------------------------------------




## Missing Outcome Data: Inverse Weighting

So far, we have not discussed the issue of missing outcome data. While the G-computation estimator is robust to arbitrary model misspecification, it is only valid if data are missing completely at random (MCAR): missingness is unrelated to either the observed or unobserved data. If this were the case, we should not see any association between the baseline covariates and missingness in the VAS opiate craving scores at the end-of-taper visit. We can assess this using a generalized additive model (GAM):

```{r Fit-GAM-Missingness-VAS-Opiate-Cravings}
vas_mar_glm <-
  mgcv::gam(
    formula = 
      is.na(vas_crave_opiates_eot) ~
      arm + stability_dose +
      s(age) + sex +
      s(vas_crave_opiates_bl) + s(arsw_score_bl) + s(cows_total_score_bl) +
      s(vas_current_withdrawal_bl) + s(vas_study_tx_help_bl) +
      uds_any_positive_bl,
    family = binomial(link = "logit"),
    data = ctn03_sim_mar
  )

summary(vas_mar_glm)
```


The GAM model shows that, all other things being equal, missingness was lower in the 7-day arm, higher in individuals on 16 or 24 mg stability doses, and was associated with some of the baseline outcome measures. In the next section, we will discuss methods that can be robust to model misspecification and do not require the MCAR assumption.
